---
title: "07-Dataframes"
author: "Tim van der Stap"
date: '2023-01-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Why dataframes? Because they can store elements of multiple types, unlike vectors and matrices. Vectors and matrices therefore can't store 'real data'. Lists are equally suboptimal and often cumbersome to work with. That being said, data frames _are_ actually a list - they combine vectors of the same length into the columns of a table. 

```{r 1, eval = TRUE}
title <- c("Star Wars", "The Empire Strikes Back", "Return of the Jedi", "The Phantom Menace", "Attack of the Clones", "Revenge of the Sith", "The Force Awakens")
year <- c(1977, 1980, 1983, 1999, 2002, 2005, 2015)
length.min <- c(121, 124, 133, 133, 142, 140, 135)
box.office.mil <- c(787, 534, 572, 1027, 657, 849, 2059)

my.data <- data.frame(title, year, length.min, box.office.mil)

# Naming columns:
names(my.data) <- c("Movie Title", "Release Year", "Length in Minutes", "Box Office")

# stringsAsFactors determines whether character vectors are converted into factors.
my.data <- data.frame(Title = title, Year = year, Length = length.min, Gross = box.office.mil,
                      stringsAsFactors = FALSE)

# Naming rows (not ideal practice)
# data.frame(..., row.names = ...)

str(my.data)
```

Exercise 16:

```{r exercise 16, eval = TRUE}
car <- c("VolkswagenTouareg", "CitroenC3", "AudiA3", "ToyotaYaris", "KiaForte", "DaciaLogan", "NissanParejo")
Year <- c(2, 3, 4, 5, 8, 9, 8)
Size <- c("medium", "small", "medium", "small", "small", "medium", "extra large")
MPG <- c(25, 47, 27, 36, 31, 2, 36)
Gearbox <- c("manual", "manual", "auto", "auto", "manual", "manual", "auto")

df <- data.frame(Year, Size, MPG, Gearbox, stringsAsFactors = TRUE)
rownames(df) <- car

# Invert the values for Gearbox:
levels(df[, "Gearbox"]) <- c("manual", "auto")
```

```{r 2, eval = TRUE}
library(tidyverse)

# General function is the read.table("file.name", sep = , header = , stringsAsFactors = ). This reads a file in table format and creates a dataframe from it. Remember that stringsAsFactors argument specifies whether all string variables have to be read in as factors or not.

my.pok <- read.table("./data_files/pokRdex_comma.csv", sep = ",", header = TRUE, stringsAsFactors = F)

# For .csv files, easier to use `read.csv("file.name", stringsAsFactors = )
my.pok <- read.csv("./data_files/pokRdex_comma.csv", stringsAsFactors = F)

# Shortcut to reading tab-delimited files: ?read.delim

# To skip a specified number of rows: read.csv(..., skip = n). Or, if you want R to read in a specific number of rows: read.csv(..., nrows = n) (header not included)
```

To share data files. Make sure to set row.names to `FALSE`, otherwise R will add a column (X, X.1 and so on) marking each row from 1 - n. But those numbers _should not_ be part of your data! This would create too many redundant columns. 

```{r 3, eval = TRUE}
write.csv(data, file = , row.names = )
write.csv(my.data, file = "MyFirstDataFrame.csv", row.names = F)

# For table-delimited table:
write.table(data, file = , row.names = )
```

Exercise 17:

```{r exercise 17, eval = TRUE}
employee <- read.csv("./data_files/employee_data.csv", stringsAsFactors = F, skip = 23, nrows = 200)
colnames(employee) <- c("Employee number", "First name", "Last name", "Birth date", "Gender", "Job title", 
                        "Salary", "From date", "To date")
write.csv(employee, file = "./data_output/EmployeeData.csv", row.names = F)

# Read it back in to make sure it's done correctly:
employee <- read.csv("./data_output/EmployeeData.csv", stringsAsFactors = F)
```

Useful ways to get a quick glimpse of what the data includes. str() gives the compact version of your data structure. summary() is a multi-purpose statistic. For numerical data is provides essential descriptive statistics. Summary() can be used on a single variable or multiple variables. 

```{r 4, eval = TRUE}
nrow(my.pok)
ncol(my.pok)

colnames(my.pok)
summary(my.pok) 
```

Subsetting dataframes. To index an element, use data[row, column]. Remember that data frames are 2D lists. Consider this when indexing and subsetting. Lists can be subsetted with the $ function, and so data frames as well. 

```{r subsetting dataframes, eval = TRUE}
library(tidyverse)
my.wars <- as.data.frame(starwars)

# Remove last three columns:
my.wars <- my.wars[ , -c(12:14)]

# Show first or last 6 observation rows:
head(my.wars)
tail(my.wars)

my.wars[3, "homeworld"]
my.wars[5, ]

head(my.wars[, "name"]) # returns a data frame with elements.
head(my.wars[, 1])
head(my.wars[["name"]]) # returns a vector!
head(my.wars$name) # returns a vector!
my.wars["name"] # returns a dataframe

my.wars[c(1:14), c("name", "homeworld", "species")] # also returns a dataframe
```

When adding observations or variables (columns). Adding columns is straightforward as these vectors have just a single element type. Rows or observations are more tricky as they likely have multiple element types (integers, characters etc) (consider the rules of coercion). Solution here is to create a small, 1 row dataframe, _where you specify the column names_.   

```{r 5, eval = TRUE} 
my.data

mark <- c(37.5, 34.75, 34.25, 0, 0, 0.75, 0)
carrie <- c(13.5, 22.75, 21.25, 0, 0, 0.5, 5.75)

# Adding columns:
my.data$MarkScreenTime <- mark
my.data$CarrieScreenTime <- carrie

# Remove a variable from a dataframe:
my.data$MarkScreenTime <- NULL
my.data$CarrieScreenTime <- NULL

my.data[["MarkScreenTime"]] <- mark
my.data[["CarrieScreenTime"]] <- carrie

# or using cbind()
my.data <- cbind(my.data, MarkScreenTime = mark, CarrieScreenTime = carrie)

# Adding observation
rogueOne <- data.frame(Title = "Rogue One", Year = 2016, Length = 133, Gross = 1051, 
                       MarkScreenTime = 0, CarrieScreenTime = 0.25)
rbind(my.data, rogueOne)

# So three ways to add columns:
data.frame$new.column
data.frame[["new.column"]]
cbind(data.frame, new.column)

# One way to add rows - column names have to match.
rbind(data.frame, new.data.frame)
```

On to Exercise 18!

```{r exercise 18, eval = TRUE}
Inspected <- c("Yes", "Yes", "No", "Yes", "No", "No", "Yes")
cars <- cbind(df, Inspected)

FordFusion <- data.frame(Year = 6, Size = "large", MPG = 27, Gearbox = "auto", Inspected = "Yes",
                         row.names = "FordFusion")
cars <- rbind(cars, FordFusion)
nrow(cars)
ncol(cars)
str(cars)

# To load in data first just do data() without assigning it.
data("diamonds")
diamonds <- data("diamonds")

data("USJudgeRatings")
judges <- USJudgeRatings
judges$AVRG <- rowMeans(judges)

# create a dataframe from that single column:
avrg <- judges["AVRG"]
```

