---
title: "10- Exploratory Data Analysis"
author: "Tim van der Stap"
date: '2023-02-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

ny <- c(1, 2, 3, 3, 5, 6, 7, 8, 9, 11)
la <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

pizza <- data_frame(ny, la)
pizza

mean(pizza$ny)
mean(pizza$la)

# Median is the middle number in an ordered(!) dataset. Not affected by outliers (unlike mean). 
median(pizza$ny)

# Mode is the value that occurs most often: R doesn't have a built-in function.
x <- table(pizza$ny)
x # This is a contingency table = shows the frequency distribution of the data. Useful for categorical data. Multiple modes are OK, but more than 3 modes defeats the purpose. 
names(x)[which(x==max(x))]

summary(pizza)

# Skewness indicates whether the observations in a dataset are concentrated to one side.

col.1 <- c(1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 4, 4, 5, 5, 7)
col.2 <- c(1, 1, 2, 2, 3, 3, 3, 4, 4, 4, 4, 4, 5, 5, 5, 6, 6, 7, 7)
col.3 <- c(1, 2, 3, 3, 4, 4, 4, 5, 5, 5, 5, 6, 6, 6, 6, 6, 6, 7, 7)

df <- data_frame(col.1, col.2, col.3)
summary(df)

# Mean > median: positive skew (or right skew). The direction of the skew is counter-intuitive. It is determined by the side the tail(!) is leaning to. So the outliers are to the right.
pl.1 <- ggplot(df, aes(x = col.1)) +
  geom_histogram(binwidth = 1, color = "white", fill = "red4") +
  theme_light() + labs(title = "Positive Skew")

# Mean = median: completely symmetrical. 
pl.2 <- ggplot(df, aes(x = col.2)) +   
  geom_histogram(binwidth = 1, color = "white", fill = "red4") +
  theme_light() + labs(title = "Zero Skew")

# Mean < median: negative (or left) skew: Tail/outliers to the left.
pl.3 <- ggplot(df, aes(x = col.3)) +
  geom_histogram(binwidth = 1, color = "white", fill = "red4") +
  theme_light() + labs(title = "Negative Skew")

# Skewness is the link between central tendency measures and probability theory.
```

Exercise 25:

```{r}
skew_1 <- read.csv("./data_files/skew_1.csv", stringsAsFactors = F)
skew_2 <- read.csv("./data_files/skew_2.csv", stringsAsFactors = F)

summary(skew_1)
# Median < Mean, so there is a positive skew (or right skew).
pl.skew_1 <- ggplot(data = skew_1, aes(x = Dataset.1)) +
  geom_histogram(binwidth = 50, color = "white", fill = "red4") + 
  theme_light()

summary(skew_2)
# Median > Mean, so there is a negative skew (or left skew)
pl.skew_2 <- ggplot(data = skew_2, aes(x = Dataset.2)) +
  geom_histogram(binwidth = 50, color = "white", fill = "red4") +
  theme_light()
```

Measures of variability: variance, standard deviation and coefficient of variation. These are univariate units of dispersion (not for interactions). 

_Variance_ measures the dispersion of a set of data points around their mean value. The closer x is to the mean, the smaller the variance. 
_Standard deviation_ = the square root of the variance.
_Coefficient of variation_ = relative standard deviation. Standard deviation relative to the mean (so standard deviation / mean). 

The _apply(...)_ family is great for running operations in a repetitive way, from element to element, but without having to write loops to do so. 
_lapply(...)_ applies the function you pass to every element in a list. You can use _lapply(...)_ to run the same code on every column in a dataframe and get a list with the results. 
_sapply(...)_ does essentially the same but returns the results in the same structure you provide (i.e. if you provide your data in a dataframe, it will provide you the results in a dataframe). 

```{r}
pizza <- as_tibble(pizza)
pizza$ny.mxn <- c(18.81, 37.62, 56.43, 56.43, 94.05, 112.86, 131.67, 150.48, 169.29, 206.91)

lapply(pizza, mean)
sapply(pizza, mean)

sapply(pizza, var)
sapply(pizza, sd)
coef.var <- sapply(pizza, sd)/sapply(pizza, mean)
coef.var
```

Multivariate measures: covariance and correlation coefficient. 


```{r}
homes <- read.csv("./data_files/landdata-states.csv", stringsAsFactors = F)
homes <- as_tibble(homes)

# Correlated:
homes %>% subset(Date == 2001.25) %>% ggplot(aes(y = Structure.Cost,
                                                 x = log(Land.Value))) +
  geom_point() + theme_light() + labs(x = "Land Value (transformed)",
                                      y = "Structure Cost (USD)",
                                      title = "Relationship between land value and structure cost.")

# Covariance gives us a sense of direction in which the two variables are moving. Positive = in the same direction. Opposite direction = negative sign. If movements are independent, covariance is 0. Comparison is meaningless. Covariance value is dependent on the scale on which the values are recorded. --> correlation coefficient: adjusts the covariance value, making it easy to interpret (standardizes covariance). 

# Correlation coefficient (will be between -1 < correlation < 1)
cor(homes$Structure.Cost, homes$Land.Value)

# Correlation coefficient = 1 = perfect positive correlation. The entire variability of one variable is explained by the other variable. Observations would form straight line in the plot. 
# Correlation coefficient = 0: no correlation. No relationship between variables, completely independent. 
# Correlation coefficient = -1: perfect negative correlation. 
cor.test(homes$Structure.Cost, homes$Land.Value)
```

Exercise 26:

```{r}
customer <- read.csv("./data_files/practical_customer.csv", stringsAsFactors = F)
customer <- as_tibble(customer)

product <- read.csv("./data_files/practical_product.csv", stringsAsFactors = F)
product <- as_tibble(product)

summary(customer)
summary(product)

# Customer.ID is character, Mortgage is character, and year of sale is year from 2004 - 2010.
plot <- ggplot(data = product, aes(x = Price)) +
  geom_histogram(binwidth = 100000) + theme_light()

# Mean is larger than the median, so there is a positive skew (tail to the right). Graph indeed shows this.
plot.2 <- ggplot(data = product, aes(y = Price,
                                     x = Area..ft..)) +
  geom_point() + theme_light()
# Positive correlation between area (sq ft) and price. 

# mean
mean(product$Price) # 281171.9

# median
median(product$Price) # 249075.7

# mode
x <- table(product$Price)
names(x)[which(x==max(x))] # 460001.256

# skewness
# Mean > median, so there is positive skewness (right skew)

# standard deviation
sd(product$Price) #89119.12

# Correlation between Price and Area:
cor(product$Price, product$Area..ft..) # 0.9510874. As expected based on scatterplot.
```
